Stack Safety as a Security Property

### 2021-02-17 main.tex
Added introduction examples

But what {\em is} this stack safety of which people speak? Precise formal
definitions are mostly lacking: rather, the notion
is commonly defined in a negative way, by enumerating the bad things that a
given enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%\bcp{Does use of unitialized variables lead to stack safety attacks?\apt{sure, via buffer overflow}}
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
Indeed, we are only aware of one quite recent attempt to frame a
positive definition, by
\citet{Skorstengaard+19}, which defines stack safety as
the conjunction of two properties: {\em local state encapsulation} and {\em
  well-bracketed control flow}.
These properties are formalized by defining a capability-based machine
that ``satisfies [stack safety] by construction.''
%
While this work represents an important step in the right direction, we believe
there is still room to craft a more abstract characterization
of stack safety that is independent of capabilities or any other specific
enforcement
technique.  In other words, we seek formal criteria that can be applied
to a mechanism for protecting stack-based low-level programs to judge
whether or not it actually ``guarantees stack safety.''
%
Our key contribution is an abstract
presentation of local state encapsulation using the technical
framework of language-based security, together with a simple
characterization of well-bracketed control flow.

The intuition behind \citeauthor{Skorstengaard+19}'s {\em
  local state encapsulation} is that callers should be protected from
callees: a callee should not be able to access or overwrite private data
in the caller's stack frame.
%
We formalize this intuition in (1) a {\em stack-confidentiality}
property, which promises that a callee's observable behavior is not affected
by memory outside its stack frame (e.g., a caller's private data or
uninitialized memory), plus (2) a {\em stack-integrity} property,
which promises that, if a function call ever returns, the caller's stack
data at the point of return will be identical to what it was before the
call.

The intuition behind \citeauthor{Skorstengaard+19}'s {\em
  well-bracketed control flow} is the need to protect the call graph
itself: a callee should pass control back to the caller only by
transferring control to the instruction
following the call; it should not be allowed to jump back into the
caller's code at an arbitrary point, either directly (via a jump
instruction) or indirectly (by corrupting stack data and
returning). Well-bracketed control flow establishes safety between
functions: each instruction is uniquely owned by some function, and
all jumps between functions are either calls that enter at a marked
entry point or returns that re-enter just after their associated call
in the standard nested call structure.

To phrase local state encapsulation as a
security property requires an explicit attacker model: a clear
understanding of who is being protected and from what. Most existing
stack protection mechanisms envisage scenarios where a callee is
influenced by external inputs to corrupt the stack in some way. To
avoid getting into the details of exactly how this happens when
formalizing our properties, we use an even stronger model: when we
state the protections that a function can rely on when it makes a
call, we assume that the callee can execute arbitrary instructions
until it returns.

A key challenge in adapting the concepts of language-based security to this
setting is that integrity and confidentiality are ``nested'' properties: not
only does a given caller need to be protected from its callee, but the callee
also needs to be protected from {\em its} callees, and so on. That is, the notion
of ``attacker'' changes at many points throughout the execution of the program.
We address
this issue by introducing a dynamically changing notion of {\em accessibility} for
the different components of the machine state and adjusting the integrity
and confidentiality levels of different stack locations and registers at
every transition (call or return) between functions.

@@

\sna{There's been some inconsistency in ``stack safety'' vs.
     ``stack-safety,'' I'm fixing on the former.}
\bcp{Be careful: When a compound phrase is used as an adjective to modify a
  following noun, it gets hyphenated; otherwise it does not.  This is
  correct grammar, not inconsistency!  :-)}
\sna{Aha, I forgot that. Corrected, and I've added it in more places.}

>>>>

But what {\em is} this stack safety of which people speak? Precise formal
definitions are mostly lacking: rather, the notion
is commonly defined in a negative way, by enumerating the bad things that a
given enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%\bcp{Does use of unitialized variables lead to stack safety attacks?\apt{sure, via buffer overflow}}
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
Indeed, we are only aware of one quite recent attempt to frame a
positive definition, by
\citet{Skorstengaard+19}, which defines stack safety as
the conjunction of two properties: {\em local state encapsulation} and {\em
  well-bracketed control flow}.
These properties are formalized by defining a capability-based machine
that ``satisfies [stack safety] by construction.''
%
While this work represents an important step in the right direction, we believe
there is still room to craft an operational characterization
of stack safety that is independent of capabilities or any other specific
enforcement technique.  In other words, we seek formal criteria that can be applied
to a mechanism for protecting stack-based low-level programs to judge
whether or not it actually ``guarantees stack safety.''
%
Our key contribution is an abstract presentation of local state encapsulation
using the technical framework of language-based security, together with a simple
characterization of well-bracketed control flow.

\begin{figure}[b]
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
main = {
  int x = 42;
  f();
  print(x);
}
\end{verbatim}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
main = {
  int x = 1;
  f();
  if(x)
    g();
  else
    h();
}
\end{verbatim}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
main = {
  int x = 1;
  f();
  x = -x;
  f();
  print(x);
}
\end{verbatim}
  \end{minipage}
  \caption{Integrity Examples}
  \label{fig:intexamples}
\end{figure}

\begin{figure}[b]
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
main = {
  int key = get_key();
  f();
  use_key(key);
}
\end{verbatim}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
main = {
  int x = get_secret();
  if(x)
    g();
  else
    h();
  f();
}
\end{verbatim}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
main = {
  int x = 0;
  int y = get_secret();
  f();
  print(x);
}
\end{verbatim}
  \end{minipage}
  \caption{Confidentiality Examples}
  \label{fig:confexamples}
\end{figure}

\paragraph*{The Use of Stack Safety}

When does the programmer rely on stack safety? The most natural example is the first
in figure \ref{fig:intexample}, in which {\tt main} calls stores 42 to a local variable {\tt x},
calls {\tt f}, and then prints {\tt x}. Here the programmer should be able to expect that
{\tt main} will print 42, but in C such an assumption may be wrong, if {\tt f} should
overwrite it's caller's stack data. To this end, our properties support a
{\em call rule} that can be used to reason that on return from {\tt f}, {\tt x} still
contains the same value. This is not only useful for the programmer: consider the second
example, where the call to {\tt h} should be unreachable and therefore subject to dead code
elimination, if only the compiler can soundly reason that {\tt x} does not change.
The property that enables the call rule is, in security terms, {\em integrity}: callers'
data is protected from modification by callees.

Alongside integrity, we have {\em confidentiality}, in which callers' data is protected
from leakage by their callees. Consider the first example in figure \ref{fig:confexamples},
in which {\tt main} recieves a key, then calls {\tt f}. If {\tt f} is an untrusted function,
it would be bad if it could just read {\tt key}. But there are more subtle examples,
such as the second one, where a secret value controls which function of {\tt g} and
{\tt h} are called prior to calling {\tt f}. Here, if {\tt g} or {\tt h} leaves behind
identifying information in unused stack space, {\tt f} could still deduce information
about the secret. So our confidentiality properties also support a {\em transparency rule},
that each function's behavior is independent of the state of its caller, other than arguments.
This also makes it easier, though not trivial, to prove idempotence: the channels through
which a callee can remember its own history are limited to those outside the stack.

Together, integrity and confidentiality comprise \citeauthor{Skorstengaard+19}'s
{\em local state encapsulation} property. But LSE alone is not sufficient to fully protect
callers from their callees. Consider the last example in figure \ref{fig:intexample},
a variant of the famous ``awkward example,'' in which {\tt f} is called twice and may not
be idempotent. It should be the case that {\tt main} prints -1, but if
{\tt f}'s first activation stores its return address somewhere accessible, the second
activation can return to before {\tt x} is negated, cause {\tt x} to be set to
1 again before a third activation that would return normally. Then {\tt main} would print 1.

Preventing this form of reentrancy attack is the intuition behind
\citeauthor{Skorstengaard+19}'s {\em well-bracketed control flow}, which protects
the call graph itself. A callee should pass control back to the caller only by
transferring control to the instruction following the call; it should not be allowed
to jump back into the caller's code at an arbitrary point, either directly (via a jump
instruction) or indirectly (by corrupting stack data and returning). It also maintains
the integrity of the stack structure; otherwise a callee could leak data in the third
example of \ref{fig:conf-example} by failing to properly restore the stack pointer.
As {\tt x} is accessed by its offset from the stack pointer, an adversary could leave
the stack pointer so that {\tt x}'s offset now points to {\tt y}, and trick {\tt main}
into printing its secret.

To phrase LSE and WBCF as security properties requires an explicit attacker model:
a clear understanding of who is being protected and from what. Most existing
stack protection mechanisms envisage scenarios where a callee is
influenced by external inputs to corrupt the stack in some way. To
avoid getting into the details of exactly how this happens when
formalizing our properties, we use an even stronger model: when we
state the protections that a function can rely on when it makes a
call, we assume that the callee can execute arbitrary instructions
until it returns.

A key challenge in adapting the concepts of language-based security to this
setting is that integrity and confidentiality are ``nested'' properties: not
only does a given caller need to be protected from its callee, but the callee
also needs to be protected from {\em its} callees, and so on. That is, the notion
of ``attacker'' changes at many points throughout the execution of the program.
We address this issue by introducing a dynamically model of {\em domains} for
the different components of the machine state, in which some stack locations are
sealed by the caller that needs to preserve them while the rest are unsealed and
available for use.

First we introduce this model for a single simple stack with no sharing between callers and
callee. Then we expand it to allow components to be passed, granting a callee one time
access to the caller's data, or shared, allowing components to be outright
exempted from the stack safety properties. Shared components could model objects whose
addresses are passed and whose security must be modeled by a separate property, such as
memory safety enforced by capabilities.

\sna{Rewriting done up to here, then skips down to property details (sections 4/5).}

@@
